---
title: 'vegan workshop - Eco-Data-Science'
author: "Caroline Owens"
date: "3 December 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

More details regarding ordination and visualization can be found [here](http://rpubs.com/an-bui/553683).

What is community data?
Rows are sites, columns are species, cell values are # occurences
Goals: 
1) describe characteristics - how speciose, how diverse
2) determine effects of environmental factors/groupings on community patterns

What is vegan?/the tidyverse?

Standard tools to import, clean, analyse, and visualize data

Today's data: birds at 210 sites with 3 landtypes (dry, riparian, mixed)
Info for each site on tree canopy structure (stems per hectare, basal area, % canopy cover, canopy height) = the environmental variables that may affect bird community structure


```{r libraries-data, message = FALSE}
# libraries
library(tidyverse)
library(vegan)

# data
# bird communities
birds <- read_csv("bird-comm.csv") %>% 
  column_to_rownames("site") #this is important for separating site names from count data

# environmental variables
env <- read_csv("env-var.csv")

# set up a "metadata" frame - will be useful for plotting later!
site_type <- env %>% 
  # selects the named columns
  select(site, landtype)
```

### How speciose are my communities?

```{r species-richness}
#vegan::specnumber counts the number of species in each site
#output is a named vector (rows in your dataframe should be named by site)
sppr <- specnumber(birds)

#simple analysis of variance to see if mean # spp is different by habitat
#stats::aov(response ~ independent variable, data = group_df)
# here, how is species richness a function of landtype 
sppr_aov <- aov(sppr ~ landtype, data = site_type)
summary(sppr_aov) #no sig diff among landtypes
```

```{r species-richness-plot}
# create dataframe for plotting
sppr_df <- sppr %>% 
  # take spp richness and convert the named vector into a dataframe
  enframe() %>% 
  # and link it with the metadata frame site_type
  full_join(site_type, by = c("name" = "site"))
#your output is a df of sites, spp numbers, and land types 
sppr_plot <- ggplot(sppr_df, aes(x = landtype, y = value, color = landtype)) +
  geom_boxplot()
sppr_plot
```

### How diverse are my communities?

Pick the function that is most relevant to your data...

```{r shannon-diversity}
# vegan::diversity() will calculate shannon (default), simpson, and fisher indices
shannondiv <- diversity(birds)
# again creates named vector 
# try simpson's diversity
simpsondiv <- diversity(birds, index = "simpson")
```



```{r shannon-diversity-plot}
shandiv_df <- shannondiv %>% 
  # convert named vector to dataframe
  enframe() %>% 
  # join with metadata: this joins shandiv_df to site_type matching shandiv_df$name to site_type$site
  full_join(., site_type, by = c("name" = "site")) %>% 
  # then calculate mean diversity for each landtype
  group_by(landtype) %>% 
  summarize(mean = mean(value))

shandiv_plot <- ggplot(shandiv_df, aes(x = landtype, y = mean, fill = landtype)) +
  geom_col()
shandiv_plot
```

### How different are my communities in species composition?

Ordination: each species is an axis along which your sites fall. Ordination compresses all these axes into 2
There are many different ways to do this math - here are three of the most basic ones.

#### perMANOVA
Permutational analysis of variance: are the centroids of my communities different?
```{r bird-permanova}
# vegan::adonis() takes the same input format as stats::aov() = response ~ IV, data = group_df
bird_perm <- adonis(birds ~ landtype, data = env)
bird_perm
```

#### PCA

There are as many PCs as there are columns in your matrix, but you can plot as many as you want... best to see what you can do with 2 or maybe 3 for best visualization

```{r bird-PCA}
# Principal components analysis = a rotation of axes
# redundancy analysis (rda) is an unconstrained ordination - the variation based ONLY on species data
# a constrained ordination would ask how env variables shape community composition
birdPCA <- rda(birds) #no env variables in here
birdPCA
# pay attention to the inertia term in output - inertia is all the variation in community composition that
# exists in your dataset

summary(birdPCA)
# this gives a big output of where species fall in the ordination (species scores and site scores) as well as
# importance of components - again components are made up just of spp abundances

pcabiplot <- biplot(birdPCA)
# this plot is not necessarily that informative
# str(pcabiplot) : you have coordinates for points and for ends of arrows

# can extract informative info from the biplot
PCAscores <- as.data.frame(pcabiplot$sites) %>% 
  bind_cols(site_type, .)

PCAvect <- as.data.frame(pcabiplot$species)


PCA_plot <- ggplot(PCAscores) +
  geom_point(aes(x = PC1, y = PC2, color = landtype)) +
  geom_segment(data = PCAvect, aes(x = 0, y = 0, xend = PC1, yend = PC2), #make segements coming out from origin
               arrow = arrow(length = unit(0.2, "cm"))) + #arrow length indicates contribution to component
              # sharp angle between arrows indicates correlation between species; 90 deg -> independence
  geom_text(data = PCAvect, aes(x = PC1, y = PC2, label = rownames(PCAvect)))

PCA_plot #expects linear response of species abundance to env variables - not always a valid assumption
```

#### NMDS

Non-metric Multidimensional Scaling

Good for nonlinear responses (eg a unimodal response)
How might communities separate according to their dissimilarity to each other?
NMDS is again collapsing variation into two axes but slightly differently than PCA

Construct dissimilarity matrix (species ~ site -> site ~ site); then use the dissimilarity as input to ordination space

default dissimilarity measure is bray-curtis distance - use the one that makes most sense for your data

```{r bird-NMDS}
#
bird_NMDS <- metaMDS(birds)

bird_NMDS

nmdsplot <- plot(bird_NMDS) #circles are sites (communities/rows), crosses are species
  #NMDS is a mapping not a rotation - axes don't matter wrt variation, etc - it's just a framework
  #for describing dissimilarity between communities

stressplot(bird_NMDS)
#stress = how far the community moves from its original state in dissimilarity
#this output is not ideal - too much stress

```

```{r NMDS-plot}
# exracting outputs
nmds_df <- as_tibble(bird_NMDS$points) %>% 
  # bind with metadata to plot
  bind_cols(site_type, .)

nmds_plot <- ggplot(nmds_df, aes(x = MDS1, y = MDS2, color = landtype, shape = landtype)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse()
nmds_plot
#all 3 habitats have different habitat structures but lots of overlap...
```

?betadisper should always be used to check assumptions of nmds (?)

##### Things to consider about stress

```{r subsampled-NMDS}
sub <- birds[sample(nrow(birds), 20), ]
subNMDS <- metaMDS(sub)
stressplot(subNMDS) #this subsample brings the stress level down to something that would be acceptable for this ordination
```

### How is community structure related to specific environmental variables?

```{r bird-CCA}
# Canonical correspondence analysis
# only shows variation in communities based on environmental variables
# like a hypothesis test for what you think is most likely to contribute to variation - don't just throw all the variables in
# good place to try several models and do model selection?
birdCCA <- cca(birds ~ canopy_height + stems_ha + big_stem_bas, data = env)
birdCCA

#again, measure of inertia (amt of variation in community composition)
#compare constrained and unconstrained inertia

```

```{r bird-CCA-plot}
ccaplot <- plot(birdCCA)
#this plot reads similar to PCA - x, y axes are the two components that describe most variation and arrows describe how communities might separate along env variables

# scaling factor is taken from structure of plot str(ccaplot) $biplot
ccavectors <- as.data.frame(ccaplot$biplot * 7.69) #scaling factor will change based on the size of your R window / plot output

#coordinates from sites and species from biplot output
site_data <- as.data.frame(ccaplot$sites) %>% 
  bind_cols(site_type, .)

species_data <- as.data.frame(ccaplot$species)

cca_plot <- ggplot(site_data) +
  geom_point(aes(x = CCA1, y = CCA2, color = landtype), shape = 19) +
  geom_segment(data = ccavectors, aes(x = 0, y = 0, xend = CCA1, yend = CCA2), arrow = arrow(length = unit(0.2, "cm"))) +
 # scale_x_continuous(limits = c(-10, 16)) + #these cut off part of the plot - comment them out to fix error
 # scale_y_continuous(limits = c(-3, 12)) +
  geom_point(data = species_data, aes(x = CCA1, y = CCA2), shape = 17, size = 2, color = "blue") +
  geom_text(data = ccavectors, aes(x = CCA1, y = CCA2, label = rownames(ccavectors)))
cca_plot

#length of arrows indicates relative importance to the ordination - here stem basal area is more important than canopy height or stems per hectare
#direction of the arrows indicates correlation between variables
#location of sites/species related to arrows is more important now: more stems/ha in dry sites, taller canopies in riparian sites
```